Implementing Capon Beamforming on the GPU for Real Time Cardiac Ultrasound Imaging

### 2013-01-24 HRUIonGPU/jrnl_article.tex
uffc

Previous work on parallel implementations of the Capon beamformer have mainly been focused around radar and passive sonar systems using custom made systolic array processors \cite{McWhirter1989, Moonen1993, Sinha2002}. For medical ultrasound, Chen et al. \cite{Chen2011a, Chen2011} implemented a Capon beamformer on the GPU that produced real apodization weights based on real data. Under this constraint the Capon beamformer is not allowed to micro-steer the main lobe, and therefore its full potential is not reached. Micro-steering of the main lobe is an important property of the Capon beamformer in order for it to improve structural details and edge definitions throughout an ultrasound image. Another aspect of using complex data is that even though the number of arithmetic operations per sample is increased we can sample at the system bandwidth, keeping the number of samples for a full cardiac imaging sector at a minimum. All results in this paper is reported with the use of baseband data. Meaning that the sampling frequency is close to the system bandwidth.

@@

Finally the Capon weight vector is formed from (\ref{eq:b}) as
\begin{align}\label{eq:w}
\vec{w}[n] = \frac{\vec{b}[n]}{\vec{1}^H\vec{b}[n]} \in \mathbb{C}^L,
\end{align}
and the beamformer output is calculated by combining all $N_L$ subarrays weighted with the same set of adaptive weights. Formally stated as
\begin{align}
z[n] &= \frac{1}{N_L}\vec{w}[n]^H \sum_{l=0}^{N_L-1} \vec{x}_l[n] \label{eq:z_mv}\\
&= 1/N_L \, \text{CONV}_m(\vec{w}[n], \vec{1}_{N_L})^H\vec{x}[n] \label{eq:z_mv2},
\end{align}
%where (this matrix is wrong)
%\begin{align}
%\mat{C} = \left(
%\begin{matrix}
%\vec{1}_1 & \vec{1}_2 & \cdots & \vec{1}_{N_L} & \vec{0}_1 & \cdots & \vec{0}_{N_L-1} & \vec{0}_{N_L}\\
%\vec{0}_{N_L-1} & \vec{0}_{N_L-2} & \cdots & \vec{0}_{0} & \vec{1}_{N_L-1} & \cdots & \vec{1}_{2} & \vec{1}_{1}\\
%\end{matrix}
%\right)^T
%\end{align}
where $\text{CONV}_m$ is a non-truncated convolution across elements.
Note how we can choose to first sum all subarrays (\ref{eq:z_mv}), or accumulate the right weighting per element (\ref{eq:z_mv2}).

In passive systems, assuming stationary statistics, the sample covariance matrix in (\ref{eq:R}) is typically averaged over a long sequence of samples\cite{Krima}. For an active, pulsed system like medical ultrasound imaging, the available data from which $\mat{\hat{R}}$ can be estimated is quite limited. Because of this, the resulting sample covariance matrix is often found to be both inaccurate, and poorly conditioned. Another challenge with applying Capon to active systems is the effect of signal cancellation, where high correlation between signal and interference favors weights that cancel out the signal \cite{Reddy1987}. In this way, the beamformer output can be too low at broadside, even when there is signal present and the distortionless-response criterion is applied.

In order to get a well-conditioned $\mat{\hat{R}}$, avoid signal cancellation, and to get DAS-like speckle, $\mat{\hat{R}}$ has to be averaged over $L\le M/2$ long subarrays and $N_K \sim T_p/T_s$ time samples \cite{Synnevag2007, Synnevag2007a}. Where, $T_p$ is the pulse length in seconds and $T_s$ is the sampling period. For baseband data, one pulse length ($1.5\lambda$) is around 3 samples ($K=1$).

@@

An SM has available a limited amount of near-core memory and registers. Each thread therefore has to be as light on memory as possible and has to perform enough instructions per used byte in order to hide memory latency. It is also important that the problem can be divided into enough threads to hide instruction latency. Global reads and writes on the GPU, in addition to CPU-GPU transfers, have high penalty, and should be minimized as much as possible. All of these constraints combined pose a real challenge as the target problem has to map on to the parallel framework both on a grid, block, thread and memory consumption level in order to benefits from parallel acceleration. In the next section, we will describe how this mapping could be done for the Capon beamforming algorithm.

@@

\section{Parallel Capon}\label{sec:meth}
If we analyze the beamformer in (\ref{eq:z_mv}) in combination with (\ref{eq:b}) and (\ref{eq:R}), we see that each calculation of a set of Capon weights is independent across the outputs $z[n]$. As shown in Fig. \ref{fig:gpulayout}, a first level of granularity is therefore to divide the image into a grid, where the output in each cell can be calculated independently of all others. Further, based on memory dependencies and to hold memory consumption per thread at a minimum, one should recognise that a group of threads should process one output in the final image. Continuing with the calculation of one weight vector, the algorithm can be broken down into three main steps that, to some extent, can be divided further into a set of parallel processes. That is, for each data vector $\vec{x}$ in the image:
\begin{align*}
%\begin{array}{l}
\text{1) } &\text{Calculate the sample covariance matrix } \mat{\hat{R}} \text{ as in } \text{(\ref{eq:diag})}\\
\text{2) } &\text{Solve the linear system of equations in (\ref{eq:b}})\\ %\vec{b} =\mat{\hat{R}}^{-1}\vec{1}_L\\
\text{3) } &\text{Calculate } \vec{w} = \vec{b}/\vec{1}_{L}^H\vec{b} \text{ and the amplitude as in (\ref{eq:z_mv})}
%\end{array}
\end{align*}
%\begin{enumerate}
%\item Calculate the sample covariance matrix $\mat{\hat{R}}$ as in (\ref{eq:R})
%\end{enumerate}
%\begin{enumerate}
%\item[2)] Solve the linear system of equations $\mat{\hat{R}}\vec{b} =\vec{ 1}$
%\end{enumerate}
%\begin{enumerate}
%\item[3)] Calculate $\vec{w} = \vec{b}/\vec{1}^H\vec{b}$ and the amplitude $\vec{w}^H\vec{x}$
%\end{enumerate}
The computational complexity of these three steps is $O(N_KN_LL^2)$, $O(L^3)$, and $O(L)$ respectively. As $M$ increases (and thereby $L$), the first two steps quickly become unmanageable, and real-time performance can only be achieved by exploring novel solutions to reduce complexity.

@@

From (\ref{eq:R}) it is obvious that calculation of one element in $\mat{\hat{R}}$ is independent of all other elements. A natural level of granularity would then be to assign one thread to each element of $\mat{\hat{R}}$ as done by \cite{Chen2011}. This will minimize global reads and writes needed per thread, but unfortunately the maximum number of threads per block puts an upper limit on the subarray length, $L_{max}=32$, since we want to avoid block-to-block synchronization.

The cubic complexity with $L$ and linear complexity with $N_K$ can be reduced by realizing that the calculation in (\ref{eq:R}) overlap both across subarrays and in time. Element $(i,j)$ of $\mat{\breve{R}}[n]$ can be calculated from element $(i-1, j-1), \text{for } i,j \in [1, L-1]$ as
\begin{align}
\mat{\breve{R}}_{i,j}[n] &=  \mat{\breve{R}}_{i-1,j-1}[n]  \nonumber \\
&+ \sum_{n'=n-K}^{n+K} (x_{i+N_L-1}[n']x_{j+N_L-1}[n']^* \nonumber \\
&- x_{i-1}[n']x_{j-1}[n']^H). \label{eq:sliding}
\end{align}
The same is true in time where
\begin{align}
\mat{\breve{R}}[n] &= \mat{\breve{R}}[n-1] + \sum_{l=0}^{N_L-1} (\vec{x}_l[n+K]\vec{x}_l[n+K]^H \nonumber \\
&- \vec{x}[n-K-1]\vec{x}[n-K-1]^H).
\end{align}
%- \mat{\breve{R}}[n-K-1] + \mat{\breve{R}}[n-K].
Both sliding across subarrays and in time can be realized simultaneously if we hold a $N_K$-average in time of the outer products $\vec{x}[n]\vec{x}[n]^H$ in memory and calculate $\mat{\hat{R}}$ averaged over subarrays from this buffer (cite JOE?). However, since $N_K$ typically is small, and we have a limited amount of memory available per thread, we focus on sliding across subarrays only. In that case, the complexity of calculating $\mat{\hat{R}}$ is reduced to $O(N_KL^2)$. Note that sliding will break independence along the sliding dimension. With our choice of calculating $\mat{\hat{R}}$, calculations are still independent across time, but not along diagonals in $\mat{\hat{R}}$. One thread is therefore used per diagonal in the upper triangle of $\mat{\hat{R}}$, and for small values of $L$, several covariance matrices are calculated per thread block (Fig. \ref{fig:gpulayout}). Maximum supported $L$ is with this approach increased to 512, however for small $L$ high $M$ configurations, memory consumption per thread will be to high to achieve full occupancy of the GPU.

@@

Diagonal loading as a constant value can be added with negligible cost. For the adaptive diagonal loading factor in (\ref{eq:diag}), the trace of $\mat{\hat{R}}$ has to be known before diagonal elements are written back to global memory. Since one row for all covariance matrices is written to global memory per kernel iteration, the diagonal elements need to be held in a local memory until the trace is accumulated. When it is found, scattered writes of the scaled diagonal elements to global memory has to be performed.

@@

\subsection{Solving Multiple Small Linear Systems}
In the literature, when it comes to solving systems of linear equations on the GPU the focus has mostly been on large systems, and GPU libraries therefore often lack a comprehensive collection of batched solvers. The reason for this is two-fold, first the GPU needs to solve thousands of small systems in order to beat the low memory latency of the CPU, second there is a range of system dimensions that does not fit well with the GPU architecture, where only a small amount of near-core memory and registers are available. Because of this, it has been proven hard to optimize a manageable set of solvers that provides speedups compared with the CPU in every cases. It is therefore not given that solving on the GPU is faster than solving on the CPU in general. However, there is two arguments for doing solving on the GPU. First the CPU, which might already runs a lot of ultrasound related algorithms, is offloaded. Second the transfer of several thousand covariance matrices from CPU-side to GPU-side is spared.

For the results presented in this paper, we have used an GPU implementation of Gauss-Jordan (GJ) elimination made by Nvidia to solve $\mat{\hat{R}}\vec{b} = \vec{1}_L$, where $\vec{b} = \mat{\hat{R}}^{-1}\vec{1}_L$. The solver has been made available for registered CUDA developers.%For smaller systems, $L < 10$, a direct solver using one thread per system is preferable. For this purpose we have implemented Cramer's rule for $2 \times 2$, $3 \times 3$ and $4 \times 4$ matrices. All solvers have been benchmarked against the latest Intel MKL libraries.

@@

\subsection{Compute Beamformer Output}
To calculate the beamformer output we have implement the formula in (\ref{eq:z_mv}) with one thread per element-weight ($L$ threads per weight vector). After the solution vector $\vec{b}$ has been placed in shared memory, one of the $L$ threads finds the sum of $\vec{b}$. Another option is to let several threads cooperate to form the sum using a binary reduction sum. However, in the following benchmarks the first approach has been used. Each of the $L$ threads then read one element of $\vec{b}$, scale it with the inverse sum just found to get the weight vector $\vec{w}[n]$.

The final output, as shown in (\ref{eq:z_mv}) and (\ref{eq:z_mv2}), can be formed by either reducing the data down to length $L$, or increase the weight vector to length $M$. Since we have decided to have $L$ threads per output, the data is reduced to length $L$ by letting thread $r$ calculate 
\begin{align}\label{eq:subsum}
\breve{x}_{r} = \sum_{k=r}^{r+N_L-1}x_{k} \text{ for } r \in [0, L-1].
\end{align}
Each thread now computes $z_r = w_r^*x_r$, and finally one thread computes and outputs $z = \frac{1}{N_L}\sum_{r=0}^{L-1} z_r$ to global memory. The last sum can also be performed using a binary reduction sum across several threads for increased speed.

\section{Beamspace Processing}\label{sec:bs}
Nilsen et al. \cite{Nilsen2009} have proposed to apply beamspace processing to further reduce the required computations for the Capon beamformer. With beamspace we mean transforming the channel data from element space to a fan of beams covering the sector illuminated by transmit beams. The transformation is written as
\begin{align}
\vec{x}_{BS}[n] &= \mat{B}\vec{x}[n], \ \ \ \mat{B} \in \mathbb{C}^{N,M}\\
[\mat{B}]_{p,q} &= \frac{1}{\sqrt{M}}e^{-j 2 \pi p q / M}.
\end{align} 
The steering vectors in $\mat{B}$, the so-called Butler matrix, determine each beam's direction in space. We can therefore remove beams where no interference is present by removing rows from $\mat{B}$.

@@

The beamspace version of the Capon beamformer is derived by replacing the covariance matrix in (\ref{eq:w}) with the beamspace covariance matrix $\mat{\hat{R}}_{BS} = f(\vec{x}_{BS}) = f(\mat{B}\vec{x})$, where $\mat{B} \in \mathbb{C}^{N_b,L}$, and $N_b$ is the number of selected beams. The transformation must be performed per subarray, yielding subarrays that no longer overlap in space. Therefore, if the transformation is applied before the covariance matrix is constructed, the amount of input data is change from $M$ to $N_LN_b$ elements per sample, however all remaining steps will get lower complexity when $N_b < L$.

@@

From (\ref{eq:sliding_bs}) it is obvious that instructions can be saved if we utilize the overlap across subarrays, with an increase in task parallel granularity (As discussed in Section \ref{sec:calcR}). With SBS we can therefore select to have one thread per beam (index $k$) that calculates this beam for all subarrays. However, since the number of beams usually is small ($N_b \ll M$) we will end up with a large memory per thread ratio. Using one thread per beam is therefore not feasible for small $N_b$ large $M$ configurations. The solution is to combine SBS with full computations. As shown in Fig. (???), a number of thread groups target different sets of subarrays which they transform using the SBS algorithm. In that way, the number of threads per sample is increased and thus the memory per thread ratio is decreased.

As mentioned the amount of data is increased from $M$ to $N_LN_b$ per sample in beamspace. This means that we no longer can exploit spatial overlap when calculating the sample covariance matrix, but since $N_b$ is small we can now allow to have one thread per element in $\mat{\hat{R}}_{BS}$. A kernel similar to the one described in \cite{Chen2011}, has therefore been implemented. For small $N_b$ values, several covariance matrices are calculated per thread block to achieve high occupancy of the SMs.

After solving $\vec{b}_{BS} = \sqrt{L}\mat{\hat{R}}_{BS}^{-1}\vec{e}_1$, the beamspace weight vector is calculated as 
\begin{align}
\vec{w}_{BS} = \frac{\vec{b}_{BS}}{\sqrt{L}\vec{e}_1\vec{b}_{BS}} \in \mathbb{C}^{N_b}
\end{align}  
and applied to the beamspace data summed across subarrays.

@@

As presented in (\ref{eq:diag}), adaptive diagonal loading is performed by loading the sample covariance matrix with the average energy per element, hence $trace\{\mat{\hat{R}}\}/L$. If adaptive diagonal loading is applied after data have been transformed into beamspace, a high percentage of the total energy is mapped to a small subset of beams. Thus the average energy per beam is higher than the average energy was per element, when beams containing little energy are removed. This means that less diagonal loading (smaller $d$) has to be applied in beamspace than in element space when $N_b < L$.

To summarize, by transforming data to beamspace we reduce the complexity of all the subsequent steps. Calculating the covariance matrix is now $O(N_KN_LN_b^2)$, the inversion step is $O(N_b^3)$, and calculating weights and the beamformer output are both $O(N_b)$. The added complexity of SBS is $O(N_bL)$.

@@

\begin{figure*}[!t]
\centerline{\subfloat[]{\includegraphics[width=3.5in]{gfx/benchmark_bar_32_40_400.png} \label{fig:benchUS32}}
\hfil
\subfloat[]{\includegraphics[width=3.5in]{gfx/benchmark_bar_64_80_400.png}%
\label{fig:benchUS64}}}
\caption{Benchmark of GPU Capon beamforming for a cardiac image covering a $70\degree$ wide $15$cm deep sector. Execution time is plotted as a function of subarray length $L$. The plot shows execution times for the three steps listed in Section \ref{sec:meth}. Calculating covariance matrices with three different choices of temporal averaging, solver, and the final beamformer sum. For larger arrays, more transmit beams have to be emitted for sufficient lateral sampling. (a) Shows the benchmark for a 32 element array. (b) Shows the benchmark for a 64 element array.}
\label{fig:bench}
\end{figure*}

\begin{figure*}[!t]
\centerline{\subfloat[]{\includegraphics[width=3.5in]{gfx/benchmark_bar_bs_M=64_Nx=80_Ny=400_Nb=3.png} \label{fig:benchBS3}}
\hfil
\subfloat[]{\includegraphics[width=3.5in]{gfx/benchmark_bar_bs_M=96_Nx=120_Ny=400_Nb=3.png}%
\label{fig:benchBS5}}}
\caption{Benchmark of GPU beamspace Capon beamforming for a cardiac image covering a $70 \degree$ and $15$cm deep sector. Execution time is plotted as a function of subarray length $L$.  If $L$ is kept constant on $M/2$ and execution time is plotted against $N_b$, the plot would be similar to the two in Fig. \ref{fig:bench}. (a) Shows the benchmark for a 64 elements array. (b) Shows the benchmark for a 96 elements array.}
\label{fig:benchBS}
\end{figure*}

>>>>

Previous work on parallel implementations of the Capon beamformer have mainly been focused around radar and passive sonar systems using custom made systolic array processors \cite{McWhirter1989, Moonen1993, Sinha2002}. For medical ultrasound, Chen et al. \cite{Chen2011a, Chen2011, Chen} implemented a Capon beamformer on the GPU that produced real apodization weights based on real data. Under this constraint the Capon beamformer is not allowed to micro-steer the main lobe, and therefore its full potential is not reached. Micro-steering of the main lobe is an important property of the Capon beamformer in order for it to improve structural details and edge definitions throughout an ultrasound image. Another aspect of using complex data is that even though the number of arithmetic operations per sample is increased we can sample at the system bandwidth, keeping the number of samples for a full cardiac imaging sector at a minimum. All results in this paper is reported with the use of baseband data. Meaning that the sampling frequency is close to the system bandwidth.

@@

Finally the Capon weight vector is formed from (\ref{eq:b}) as
\begin{align}\label{eq:w}
\vec{w}[n] = \frac{\vec{b}[n]}{\vec{1}^H\vec{b}[n]} \in \mathbb{C}^L,
\end{align}
and the beamformer output is calculated by combining all $N_L$ subarrays weighted with the same set of adaptive weights. Formally stated as
\begin{align}
z[n] &= \frac{1}{N_L}\vec{w}[n]^H \sum_{l=0}^{N_L-1} \vec{x}_l[n] \label{eq:z_mv}\\
&= 1/N_L \, \text{CONV}_m(\vec{w}[n], \vec{1}_{N_L})^H\vec{x}[n] \label{eq:z_mv2},
\end{align}
%where (this matrix is wrong)
%\begin{align}
%\mat{C} = \left(
%\begin{matrix}
%\vec{1}_1 & \vec{1}_2 & \cdots & \vec{1}_{N_L} & \vec{0}_1 & \cdots & \vec{0}_{N_L-1} & \vec{0}_{N_L}\\
%\vec{0}_{N_L-1} & \vec{0}_{N_L-2} & \cdots & \vec{0}_{0} & \vec{1}_{N_L-1} & \cdots & \vec{1}_{2} & \vec{1}_{1}\\
%\end{matrix}
%\right)^T
%\end{align}
where $\text{CONV}_m$ is a non-truncated convolution across elements.
Note how we can either choose to sum all subarrays (\ref{eq:z_mv}), or accumulate the right weighting per element (\ref{eq:z_mv2}).

In passive systems, assuming stationary statistics, the sample covariance matrix in (\ref{eq:R}) is typically averaged over a long sequence of samples\cite{Krima}. For an active, pulsed system like medical ultrasound imaging, the available data from which $\mat{\hat{R}}$ can be estimated is limited. Because of this, the resulting sample covariance matrix is often found to be both inaccurate, and poorly conditioned. Another challenge with applying Capon to active systems is the effect of signal cancellation, where high correlation between signal and interference favors weights that cancel out the signal \cite{Reddy1987}. In this way, the beamformer output can be too low at broadside, even when there is signal present and the distortionless-response criterion is applied.

In order to get a well-conditioned $\mat{\hat{R}}$, avoid signal cancellation, and to get DAS-like speckle, $\mat{\hat{R}}$ has to be averaged over $L\le M/2$ long subarrays and $N_K \sim T_p/T_s$ time samples \cite{Synnevag2007, Synnevag2007a}, where, $T_p$ is the pulse length in seconds and $T_s$ is the sampling period. For baseband data, one pulse length ($1.5\lambda$) is around 3 samples ($K=1$).

@@

An SM has available a limited amount of near-core memory and registers. Each thread therefore has to be as light on memory as possible and has to perform enough instructions per used byte in order to hide memory latency. It is also important that the problem can be divided into enough threads to hide instruction latency. Global reads and writes on the GPU, in addition to CPU-GPU transfers, have high penalty, and should be minimized as much as possible. All of these constraints combined pose a real challenge as the target problem has to map on to the GPU framework both on a grid, block, thread and memory consumption level in order to benefits from parallel acceleration. In the next section, we will describe how this mapping could be done for the Capon beamforming algorithm.

@@

\section{Parallel Capon}\label{sec:meth}
If we analyze the beamformer in (\ref{eq:z_mv}) in combination with (\ref{eq:b}) and (\ref{eq:R}), we see that each calculation of a set of Capon weights is independent across the outputs $z[n]$. As shown in Fig. \ref{fig:gpulayout}, a first level of granularity is therefore to divide the image into a grid, where the output in each cell can be calculated independently of all others. Further, based on memory dependencies and to hold memory consumption per thread at a minimum, one should recognise that a group of threads should process one output in the final image. Continuing with the calculation of one weight vector, the algorithm can be broken down into three main steps that, to some extent, can be divided further into a set of parallel processes. That is, for each data vector $\vec{x}$ in the image:
\begin{align*}
%\begin{array}{l}
\text{1) } &\text{Calculate the sample covariance matrix } \mat{\hat{R}} \text{ as in } \text{(\ref{eq:diag})}\\
\text{2) } &\text{Solve the linear system of equations as in (\ref{eq:b}})\\ %\vec{b} =\mat{\hat{R}}^{-1}\vec{1}_L\\
\text{3) } &\text{Calculate } \vec{w} = \vec{b}/\vec{1}_{L}^H\vec{b} \text{ and the amplitude as in (\ref{eq:z_mv})}
%\end{array}
\end{align*}
%\begin{enumerate}
%\item Calculate the sample covariance matrix $\mat{\hat{R}}$ as in (\ref{eq:R})
%\end{enumerate}
%\begin{enumerate}
%\item[2)] Solve the linear system of equations $\mat{\hat{R}}\vec{b} =\vec{ 1}$
%\end{enumerate}
%\begin{enumerate}
%\item[3)] Calculate $\vec{w} = \vec{b}/\vec{1}^H\vec{b}$ and the amplitude $\vec{w}^H\vec{x}$
%\end{enumerate}
The computational complexity of these three steps is $O(N_KN_LL^2)$, $O(L^3)$, and $O(L)$ respectively. As $M$ increases (and thereby $L$), the first two steps quickly become unmanageable, and real-time performance can only be achieved by exploring novel solutions to reduce complexity.

@@

From (\ref{eq:R}) it is obvious that calculating one element of $\mat{\hat{R}}$ is independent of all other elements. A natural level of granularity would then be to assign one thread to each element of $\mat{\hat{R}}$ \cite{Chen2011}. This will minimize global reads and writes needed per thread, but unfortunately the maximum number of threads per block puts an upper limit on the subarray length, $L_{max}=32$, since we want to avoid block-to-block synchronization. Since arrays of 96-128 elements are common in medical ultrasound imaging an implementation not limited by $L_max=32$ is preferable.

To reach this goal we start by showing how the cubic complexity with $L$ and linear complexity with $N_K$ can be reduced by realizing that calculations in (\ref{eq:R}) overlap both across subarrays and in time. Element $(i,j)$ of $\mat{\breve{R}}[n]$ can be calculated from element $(i-1, j-1), \text{for } i,j \in [1, L-1]$ as
\begin{align}
\mat{\breve{R}}_{i,j}[n] &=  \mat{\breve{R}}_{i-1,j-1}[n]  \nonumber \\
&+ \sum_{n'=n-K}^{n+K} (x_{i+N_L-1}[n']x_{j+N_L-1}[n']^* \nonumber \\
&- x_{i-1}[n']x_{j-1}[n']^H). \label{eq:sliding}
\end{align}
The same is true in time where
\begin{align}
\mat{\breve{R}}[n] &= \mat{\breve{R}}[n-1] + \sum_{l=0}^{N_L-1} (\vec{x}_l[n+K]\vec{x}_l[n+K]^H \nonumber \\
&- \vec{x}[n-K-1]\vec{x}[n-K-1]^H).
\end{align}
%- \mat{\breve{R}}[n-K-1] + \mat{\breve{R}}[n-K].
Both sliding across subarrays and in time can be realized simultaneously if we hold a $N_K$-average in time of the outer products $\vec{x}[n]\vec{x}[n]^H$ in memory and calculate $\mat{\hat{R}}$ averaged over subarrays from this buffer (cite JOE?). However, since $N_K$ typically is small, and we have a limited amount of memory available per thread, we focus on sliding across subarrays only. In that case, the complexity of calculating $\mat{\hat{R}}$ is reduced to $O(N_KL^2)$. Note that sliding will break independence along the sliding dimension. With our choice of calculating $\mat{\hat{R}}$, calculations are still independent across time, but not along diagonals in $\mat{\hat{R}}$. One thread is therefore used per diagonal in the upper triangle of $\mat{\hat{R}}$, and for small values of $L$, several covariance matrices are calculated per thread block (Fig. \ref{fig:gpulayout}). Maximum supported $L$ is with this approach increased to 512, however for small $L$ high $M$ configurations, memory consumption per thread will be to high to achieve full occupancy of the GPU.

@@

Diagonal loading as a constant value can be added with negligible cost. For the adaptive diagonal loading factor in (\ref{eq:diag}), the trace of $\mat{\hat{R}}$ has to be known before diagonal elements are written back to global memory. Since one row for all covariance matrices is written to global memory per kernel iteration, the diagonal elements need to be held in a local memory until the trace is accumulated. When that is done, scattered writes of the scaled diagonal elements to global memory has to be performed.

@@

\subsection{Solving Multiple Small Linear Systems}
In the literature, when it comes to solving systems of linear equations on the GPU the focus has mostly been on large systems, and GPU libraries therefore often lack a comprehensive collection of batched solvers. The reason for this is two-fold, first the GPU needs to solve thousands of small systems in order to beat the low memory latency of the CPU, second there is a range of system dimensions that does not fit well with the GPU architecture, where only a small amount of near-core memory and registers are available. Because of this, it has been proven hard to optimize a manageable set of solvers that provides speedups compared with the CPU in every cases. It is therefore not given that solving on the GPU is faster than solving on the CPU in general. However, there is two arguments for doing solving on the GPU. First the CPU, which might already runs a lot of ultrasound related algorithms, is offloaded. Second the transfer of several thousand covariance matrices over the PCI-Express buss, from CPU-side to GPU-side, is spared.

For the results presented in this paper, we have used a GPU implementation of batched Gauss-Jordan (GJ) elimination made by Nvidia to solve $\mat{\hat{R}}\vec{b} = \vec{1}_L$, where $\vec{b} = \mat{\hat{R}}^{-1}\vec{1}_L$ for all pixels at once. The solver has been made available for registered CUDA developers.%For smaller systems, $L < 10$, a direct solver using one thread per system is preferable. For this purpose we have implemented Cramer's rule for $2 \times 2$, $3 \times 3$ and $4 \times 4$ matrices. All solvers have been benchmarked against the latest Intel MKL libraries.

@@

\subsection{Compute Beamformer Output}
To calculate the beamformer output, equation (\ref{eq:z_mv}) has been implemented using one thread per element-weight ($L$ threads per weight vector). After the solution vector $\vec{b}$ has been placed in shared memory, one of the $L$ threads finds the sum of $\vec{b}$. 
%Another option is to let several threads cooperate to form the sum using a binary reduction sum. However, in the following benchmarks the first approach has been used. 
Each of the $L$ threads then read one element of $\vec{b}$, scale it with the inverse sum just found to get the weight vector $\vec{w}[n]$.

The final output, as shown in (\ref{eq:z_mv}) and (\ref{eq:z_mv2}), can be formed by either reducing the data down to length $L$, or increase the weight vector to length $M$. Since the design is $L$ threads per output, the data is reduced to length $L$ by letting thread $r$ calculate 
\begin{align}\label{eq:subsum}
\breve{x}_{r} = \sum_{k=r}^{r+N_L-1}x_{k} \text{ for } r \in [0, L-1].
\end{align}
Each thread now computes $z_r = w_r^*x_r$, and finally one thread computes and outputs $z = \frac{1}{N_L}\sum_{r=0}^{L-1} z_r$ to global memory. %The last sum can also be performed using a binary reduction sum across several threads for increased speed.

\section{Beamspace Processing}\label{sec:bs}
Nilsen et al. \cite{Nilsen2009} have proposed to apply beamspace processing to further reduce the required computations of Capon beamformer. With beamspace we mean transforming the channel data from element space to a fan of beams covering the sector illuminated by transmit beams. The transformation is formally stated as
\begin{align}
\vec{x}_{BS}[n] &= \mat{B}\vec{x}[n], \ \ \ \mat{B} \in \mathbb{C}^{N,M}\\
[\mat{B}]_{p,q} &= \frac{1}{\sqrt{M}}e^{-j 2 \pi p q / M}.
\end{align} 
The steering vectors in $\mat{B}$, the so-called Butler matrix, determine each beam's direction in space. We can therefore remove beams where no interference is present by simply removing rows from $\mat{B}$.

@@

The beamspace version of the Capon beamformer is derived by replacing the covariance matrix in (\ref{eq:w}) with the beamspace covariance matrix 
\begin{align}
\mat{\hat{R}}_{BS} = f(\vec{x}_{BS}) = f(\mat{B}\vec{x}), 
\end{align}
where $\mat{B} \in \mathbb{C}^{N_b,L}$, and $N_b$ is the number of selected beams. The transformation must be performed per subarray, yielding subarrays that no longer overlap in space. Therefore, if the transformation is applied before the covariance matrix is constructed, the amount of input data is change from $M$ to $N_LN_b$ elements per sample, however all remaining steps will get lower complexity when $N_b < L$.

@@

Equation (\ref{eq:sliding_bs}) shows that instructions are saved if we utilize the overlap across subarrays, but with an increase in task parallel granularity (As discussed in Section \ref{sec:calcR}). The most obvious is to use one thread per beam (index $k$) that calculates this beam for all subarrays. However, since the number of selected beams usually is small ($N_b \ll M$) we will end up with a large memory per thread ratio. Using one thread per beam is therefore not feasible for small $N_b$ large $M$ configurations. The solution is to combine SBS with full computations. As shown in Fig. (???), a number of thread groups target different sets of subarrays which they transform using the SBS algorithm. In that way, the number of threads per sample is increased and thus the memory per thread ratio is decreased.

As mentioned the amount of channel data is increased from $M$ to $N_LN_b$ per sample in beamspace. This means that we no longer can exploit spatial overlap when calculating the sample covariance matrix, but since $N_b$ is small we can now allow to have one thread per element in $\mat{\hat{R}}_{BS}$. A kernel similar to the one described in \cite{Chen2011} has therefore been implemented. In addition, for small $N_b$ values, several covariance matrices are calculated per thread block to achieve high occupancy of the GPU.

After solving $\vec{b}_{BS} = \sqrt{L}\mat{\hat{R}}_{BS}^{-1}\vec{e}_1$, the beamspace weight vector is calculated as 
\begin{align}
\vec{w}_{BS} = \frac{\vec{b}_{BS}}{\sqrt{L}\vec{e}_1\vec{b}_{BS}} \in \mathbb{C}^{N_b}
\end{align}  
and applied to the beamspace data summed across subarrays,
\begin{align}
z[n] = \vec{w}_{BS}^H\sum_{l=0}^{L-1}\vec{x}_{BS,l}. 
\end{align}

@@

As presented in (\ref{eq:diag}), adaptive diagonal loading is performed by loading the sample covariance matrix with the average energy per element, hence $trace\{\mat{\hat{R}}\}/L$. When data are transformed into beamspace, a high percentage of the total energy is mapped to a small subset of beams. Thus the average energy per beam is higher than the average energy was per element, when beams containing little energy are removed. This means that less diagonal loading (smaller $d$) has to be applied in beamspace than in element space when $N_b < L$.

To summarize, by transforming data to beamspace we reduce the complexity of all the subsequent steps. Calculating the covariance matrix is now $O(N_KN_LN_b^2)$, the inversion step is $O(N_b^3)$, and calculating weights and the beamformer output are both $O(N_b)$. The added complexity of SBS is $O(N_bL)$. To maximize performance, large adjustments to the processing pipeline described in Section \ref{sec:meth} is required.

@@

\begin{figure*}[!t]
\centerline{\subfloat[]{\includegraphics[width=3.5in]{gfx/benchmark_bar_32_40_400.png} \label{fig:benchUS32}}
\hfil
\subfloat[]{\includegraphics[width=3.5in]{gfx/benchmark_bar_64_80_400.png}%
\label{fig:benchUS64}}}
\caption{Benchmark of GPU Capon beamforming for a cardiac image covering a $70\degree$ wide $15$cm deep sector. Execution time is plotted as a function of subarray length $L$. The plot shows execution times for the three steps listed in Section \ref{sec:meth}. Calculating covariance matrices with three different choices of temporal averaging, solver, and the final beamformer sum. For larger arrays, more transmit beams have to be emitted for sufficient lateral sampling. (a)  32 element array. (b)  64 element array.}
\label{fig:bench}
\end{figure*}

\begin{figure*}[!t]
\centerline{\subfloat[]{\includegraphics[width=3.5in]{gfx/benchmark_bar_bs_M=64_Nx=80_Ny=400_Nb=3.png} \label{fig:benchBS3}}
\hfil
\subfloat[]{\includegraphics[width=3.5in]{gfx/benchmark_bar_bs_M=96_Nx=120_Ny=400_Nb=3.png}%
\label{fig:benchBS5}}}
\caption{Benchmark of GPU beamspace Capon beamforming for a cardiac image covering a $70 \degree$ and $15$cm deep sector. Execution time is plotted as a function of subarray length $L$.  If $L$ is kept constant on $M/2$ and execution time is plotted against $N_b$, the plot would be similar to the two in Fig. \ref{fig:bench}. (a) 64 elements array. (b) 96 elements array.}
\label{fig:benchBS}
\end{figure*}
In this section we present benchmarks of the presented implementations performed on a Nvidia Quadro 6000 graphics card (14 SMs, 1 TFlops theoretical single precision, 6 GB global memory, and 144 GB/s memory bandwidth). Problem sizes are calculated based on 32, 64, and 96 element phased arrays with a 2.5 MHz center frequency. Assuming a system bandwidth of 80\% we need a complex sampling frequency of 2 MHz, and when imaging down to a depth of 15 cm we end up with approximately 400 samples per range line. To correctly sample a $70\degree$ sector with 32, 64 and 96 element arrays ($\lambda/2$ pitch), we need at least 40, 80 and 120 beams respectively. The presented execution times do not include transfer of data from CPU-side to the GPU, since data are assumed to be pre-delayed and available in GPU memory. What is included are all calculations, memory transfers from global to near-core GPU memory, and near core memory access.
